---
title: "EDA Notes and Examples"
subtitle: "EDE Fall 2024"
author: "Emily Yang"
date: "2024-10-22"
output: pdf_document
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Shortcuts

-   add code chunk: Ctrl + Alt + I
-   pipe: Ctrl + Shift + M
-   assignment: Alt + -

## r code chunk settings

-   Echo = F: only show the output, not the code
-   results = "hide", message=FALSE, warning=FALSE: code will run, output won't show
-   eval = F: code won't run
-   include = F: run the code but do not show the code
-   eval = F, include = F: never run and show the code in output
-   fig.height = 3, fig.width = 5: set absolute size of figure
-   fig.dim: set both (e.g. fig.dim = c(5,3))
-   fig.cap = "Figure 1": Figure caption
-   out.width.out.height: scale figs in output docs (e.g. out.width = 80% is setting at 80% of page width)
-   fig.align: set alignment

## Import libraries

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(knitr)
```

## Import data

```{r}
#Import EPA data (from the processed_KEY folder) & fix dates
epa_data <- read.csv(
  here("Data","Processed_KEY","EPAair_O3_PM25_NC1819_Processed.csv"),
  stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))
```

## Creating tables

General markdown table:

| Item     |                               Value |
|:---------|------------------------------------:|
| Filename | EPAair_O3_PM25_NC1819_Processed.csv |
| Date     |                           2018-2019 |
| Source   |        EPA Air Quality System (AQS) |

: EPA Air Quality

```{r}
#Compute mean PM2.5 concentration for each county, sorted by value
summary_PM25 <- epa_data %>% 
  drop_na(PM2.5) %>% 
  group_by(COUNTY) %>% 
  summarize(mean_PM2.5 = mean(PM2.5)) %>% 
  arrange(mean_PM2.5)

#Use the `kable` function to produce the 2 tables in markdown format
knitr::kable(summary_PM25, 
             caption = "Mean Particulates (2.5mm)", 
             col.names = c("County", "$\\mu g/m^3$"))
```

## Creating plots

```{r}
# Boxplot for ozone levels by month in 2018
Ozone_2018 <- epa_data %>% 
  filter(Year == 2018) %>% 
  drop_na(Ozone)

Ozone_2018$Month <- as.numeric(Ozone_2018$Month)
Ozone_2018$Month <- factor(Ozone_2018$Month,
                           levels = 1:12,
                           labels = month.abb)

Ozone_2018_plot <- 
  ggplot(Ozone_2018, aes(x = Month, y = Ozone)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Month",
       y = "Ozone Levels (ppm(v))",
       title = "Ozone by Month in 2018") +
  theme(plot.title = element_text(hjust = 0.5))

print(Ozone_2018_plot)

```

## R Markdown

(Also refer to Markdown cheat sheet)

-   Include toc (table of content) and subtitle at the beginning of the report
-   toc for figures:  + listoffigures
-   inset a new page:  + newpage
-   insert tables, images, links, format bullet points
    -   Help \> Markdown Quick Reference
    -   [my link] (<http://rmarkdown.rstudio.com>)
